\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-03-18 docs/parco13/abstract.tex
Change some FV wording, add references

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, it was realized early that it helps to aggregate degrees of
freedom (DOF) at the element level, as has been done with high-order
spectral element \cite{TufoFischer99}
%finite volume (element?)
or discontinuous Galerkin \cite{HesthavenWarburton02}
methods.  GPU implementations of the latter have been proposed recently
\cite{KlocknerWarburtonBridgeEtAl09, BursteddeGhattasGurnisEtAl10}.
The finite volume method has typically been implemented using a single degree
of freedom per cell on
% either
structured
\cite{ppm, clawpack, weno}
or unstructured meshes
\cite{ddfv, ausm+, openfoam},
where higher order methods have been constructed by widening the stencil
\cite{needonehere}.
% \comment{Are you looking for
%   references to parallel AMR codes?  Or more general AMR codes based
%   on the FVM?}  \todo{No, I was looking for unstructured FV
%   references.  Would you have any, and can we add the block-AMR
%   references later?} \comment{But most finite volume methods
%   (structured or unstructured) only have a single degree of freedom
%   per mesh cell (``element''?)  Only recently have people increased
%   the order by extending the stencil (without increase the DOFs per
%   mesh cell).}
% CB: Can you fix some of the above citations?

@@

To facilitate hardware acceleration for parallel dynamic AMR, we
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  In our work, we go
beyond the traditional high-order element and define each element to
be a dense compute patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989,
% be-le:1991 (cited below)
ColellaGravesKeenEtAl07,
BerzinsLuitjensMengEtAl10} except that the
patches are not overlapping,
% (except in ghost cell regions),
% CB: we comment on this later in Section 4.
which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The \clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw}.

@@

For hyperbolic problems, we integrate the solution on a a single uniform
patch, containing $m^d$ cells,
using the wave propagation algorithm described by R. J. LeVeque and
implemented in \clawpack \cite{le:2002, clawpack}.  We assume a
single degree of freedom per cell and reconstruct a piecewise
constant solution to obtain left and right states at cell
edges.  At each edge, we solve Riemann problems to obtain left and
right going waves propagating at speeds determined from the solution
to the Riemann problem.  For scalar advection, the speed of each wave
is simply the local advection speed at the cell interface.  For non-linear
problems and systems, an approximate Riemann solver, such as a Roe solver
\cite{roesolver}, is typically used.  Wave limiters are used to surpress
spurious oscillations at sharp gradients in the flow.

Data exchanges between neighboring patches are done via layers of
ghost cells extending the dimensions along the edges of each patch.
The interior edge values of a given patch overlap the ghost cell
region of a neighboring patch.  For the second order wave propagation
algorithm, two layers of ghost cells are sufficient.  This implies that
one layer of ghost patches is sufficient for $m \ge 4$.  Neighboring
patches at the same level of refinement simply copy their interior
edge values into a neighbors' ghost cells.  Neighboring fine grid
patches average their interior edge data to a coarser neighbor's ghost
cell values.  And neighboring coarse grid patches interpolate data
from their interior edge cells to their fine grid neighbor's ghost
cell values.  To avoid loss of conservation and the creation of
spurious extrema, we use a standard conservative, limited
interpolation scheme to interpolate values from the coarse grid to
fine grid coarse cells \cite{amrclaw, chombo}.  When subcycling, time accurate
data between coarse grids is used to fill in ghost cells for fine grids.
As mentioned in the previous section, this procedure can be extended
transparently to distributed parallelism by defining an abstract exchange
routine for ghost patch data.

>>>>

Consequently, it is not clear a priori how to accelerate PDE solvers
that use adaptive mesh refinement.
%, especially when working with
%unstructured meshes \comment{This sounds we are now going to talk
%  about unstructured meshes.}.
% CB: Good point, took it out.
Of course, it was realized early that it helps to aggregate degrees of
freedom (DOF) at the element level, as has been done with high-order
spectral element \cite{TufoFischer99}
%finite volume (element?)
or discontinuous Galerkin \cite{HesthavenWarburton02}
methods.  GPU implementations of the latter have been proposed recently
\cite{KlocknerWarburtonBridgeEtAl09, BursteddeGhattasGurnisEtAl10}.
The finite volume method has typically been implemented using a single degree
of freedom per cell on
% either
structured
\cite{ppm, clawpack}
or unstructured meshes
\cite{ddfv, ausm+, openfoam},
where higher order methods have been constructed by widening the stencil, for instance in WENO methods
\cite{Shu:2009bi}.
% \comment{Are you looking for
%   references to parallel AMR codes?  Or more general AMR codes based
%   on the FVM?}  \todo{No, I was looking for unstructured FV
%   references.  Would you have any, and can we add the block-AMR
%   references later?} \comment{But most finite volume methods
%   (structured or unstructured) only have a single degree of freedom
%   per mesh cell (``element''?)  Only recently have people increased
%   the order by extending the stencil (without increase the DOFs per
%   mesh cell).}
% CB: Can you fix some of the above citations?

@@

To facilitate hardware acceleration for parallel dynamic AMR, we
build upon the forest-of-octrees paradigm because of its low overhead
and proven scalability \cite{BursteddeWilcoxGhattas11}.  This approach
identifies each octree leaf with a mesh element.  In our work, we go
beyond the traditional high-order element and define each element to
be a dense compute patch with $m^d$ DOFs.  In fact, this approach
resembles block-structured AMR \cite{be-ol:1984, be-co:1989,
% be-le:1991 (cited below)
ColellaGravesKeenEtAl07,
BerzinsLuitjensMengEtAl10} except that the
patches are not overlapping,
% (except in ghost cell regions),
% CB: we comment on this later in Section 4.
which enables us to capitalize on our
previous experience with scalable FE solvers for PDEs
\cite{BursteddeStadlerAlisicEtAl13}.  The \clawpack software
\cite{LeVeque97} provides a popular implementation of such a patch.
It has been designed to solve hyperbolic conservation laws
% on a uniform compute patch
and successfully used in the context of block-structured AMR
\cite{be-le:1991, amrclaw, Berger:2011du}.

@@

For hyperbolic problems, we integrate the solution on a a single uniform patch,
containing $m^d$ cells, using the wave propagation algorithm described by R. J.
LeVeque \cite{LeVeque:1997eg} and implemented in \clawpack \cite{le:2002,
clawpack}.  We assume a single degree of freedom per cell and reconstruct a
piecewise constant solution to obtain left and right states at cell edges.  At
each edge, we solve Riemann problems to obtain left and right going waves
propagating at speeds determined from the solution to the Riemann problem.  For
scalar advection, the speed of each wave is simply the local advection speed at
the cell interface.  For non- linear problems and systems, an approximate
Riemann solver, such as a Roe solver \cite{roesolver}, is typically used.  Since
much of the physics of an application can be contained in the Riemann solver,
\forestclaw adopts \clawpack's interface to Riemann solvers effectively allowing
problems solved with \clawpack to be solvable in \forestclaw.  In order to
achieve second order accuracy wave limiters are used to suppress spurious
oscillations at sharp gradients in the flow.

Data exchanges between neighboring patches are done via layers of
ghost cells extending the dimensions along the edges of each patch.
The interior edge values of a given patch overlap the ghost cell
region of a neighboring patch.  For the second order wave propagation
algorithm, two layers of ghost cells are sufficient.  This implies that
one layer of ghost patches is sufficient for $m \ge 4$.  Neighboring
patches at the same level of refinement simply copy their interior
edge values into a neighbors' ghost cells.  Neighboring fine grid
patches average their interior edge data to a coarser neighbor's ghost
cell values.  And neighboring coarse grid patches interpolate data
from their interior edge cells to their fine grid neighbor's ghost
cell values.  To avoid loss of conservation and the creation of
spurious extrema, we use a standard conservative, limited
interpolation scheme to interpolate values from the coarse grid to
fine grid coarse cells \cite{amrclaw, chombo}.  When sub-cycling, time accurate
data between coarse grids is used to fill in ghost cells for fine grids.
As mentioned in the previous section, this procedure can be extended
transparently to distributed parallelism by defining an abstract exchange
routine for ghost patch data.
